
\chapter{List of publications}\label{chap:publications}

%\addcontentsline{toc}{chapter}{\bibname}
\section*{Published papers}
\begin{itemize}
	\item P. Klenovský, D. Hemzal, P. Steindl, M. Zíková, V. Křápek and J. Humlíček, \textit{Polarization anisotropy of the emission from type-II quantum dots}, Phys. Rev. B \textbf{92}, 241302 (2015).
	\item P. Klenovský, P. Steindl, D. Geffroy, \textit{Excitonic structure and pumping power dependent emission blue-shift of type-II quantum dots}, Scientific Reports \textbf{7}, 45568 (2017).
\end{itemize}

\section*{Poster presentations}
\begin{itemize}
	\item[1.]  P. Klenovský, D. Hemzal, P. Steindl, M. Zíková, V. Křápek and J. Humlíček, \textit{Polarization anisotropy of the emission from type-II quantum dots}, TWINFUSYON summerschool Advanced School on Modelling and Statistics for Biology, Biochemistry and Biosensing, Linz (Austria), 2016. 
	
	
	\item[2.] P. Steindl, P. Klenovský, \textit{Excitonic structure and pumping power dependent emission blue-shift of type-II quantum dots}, TWINFUSYON winterschool New Frontiers in 2D materials: Approaches \& Applications, Villard-de-Lans (France), 2017.
	
\item[3.] P. Steindl, P. Klenovský, D. Geffroy, \textit{Multi-excitonic structure of type-II quantum dots}, 81. Jahrestagung der DPG und DPG-Frühjahrstagung, Dresden (Germany), 2017. 

\item[4.] P. Steindl, P. Klenovský, D. Geffroy, \textit{Multi-excitonic structure of type-II quantum dots}, 16$^{\mathrm{th}}$ IUVSTA International Summer School on Physics at Nanoscale, Milovy (Czech Republic), 2017.	

\item[5.] P. Steindl, P. Klenovský, D. Geffroy, \textit{Multi-excitonic structure of type-II quantum dots}, 46$^{\mathrm{th}}$ International School \& Conference on the Physics of Semiconductors ``Jaszowiec 2017'', Szczyrk (Poland), 2017. 

\item[6.] P. Steindl, P. Klenovský, E.~M. Sala, B. Alén, D. Bimberg, \textit{Identification of individual transitions in InGaAsSb/GaP quantum dot by power and temperature dependent photoluminescence}, 82. Jahrestagung der DPG und DPG-Frühjahrstagung, Berlin (Germany), 2018.


\item[7.] P. Steindl, P. Klenovský, E.~M. Sala, B. Alén, D. Bimberg, \textit{Recombination dynamics of InGaAsSb/GaP quantum dots}, 82. Jahrestagung der DPG und DPG-Frühjahrstagung, Berlin (Germany), 2018.




\end{itemize}
\newpage 